Constructing a universe for the setoid model

### 2020-06-25 paper.tex
More paper writing

In Section~\label{sec:metatheory}...

@@

This section describes $MLTT^{Prop}$, the metatheory in which the model
construction described in the following sections takes place. One aspect of the
metatheory that we pay attention to is it's minimality, as we want to keep it as
simple as possible. We want to particularly avoid having to assume strong
definition schemas that go beyond inductive families (or, semantically, what can
be modeled with W types.)

@@

\section{Setoid model}

@@

By \emph{setoid model} we intend a class of models of (intensional) type theory
where contexts/closed types are interpreted as setoids, i.e. sets with an
equivalence relation, and dependent types are interpreted as dependent/indexed
setoids.
%
A setoid model was first given by M. Hofmann in \cite{hofmann-ext-int} in order
to give a semantics to extensionality principles such as functional and
propositional extensionality. Here we consider a similar model construction, due
to T. Altenkirch \cite{setoid-model}.
%
The peculiarity of this model is that it is presented in a type theoretic and
intensional metatheory, in our case $MLTT^{Prop}$.
%
The use of strict propositions to define the setoid equivalence relations makes
this model \emph{strict}. A strict model is one where every equation holds
definitionally. The setoid model thus defined validates extensionality
principles like funext, proext, and quotient types. As a consequence, it
provides a way to bootstrap and explain extensionality by implementing it in a
core intensional type theory.

As any model of Type Theory, the setoid model can be framed categorically as a
category with families (CwF) with extra structure for the various type and term
formers. The core structure can be defined as the following signature:

@@

Contexts are given by setoids, that is types together with an equivalence relation.

\[
\infer{
  \begin{aligned}
    & | \GG | : \mType \\
    & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
    & ...
  \end{aligned}
}{
  \GG : \Con
}
\]

Types in a context $\GG$ are given by setoid morphisms from $\GG$ to the
(strict) category of setoids. The definition can be unfolded as follows:

\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & \textsf{coe} : \{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \dots
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

This definition of types in the setoid model is different from the original one
in [cite LICS 99 paper], but it is equivalent to it [cite].
%
Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
fibration $\GG . A \to \GG$. Note that we only need to include components for
the functorial action on objects and morphisms, since the functor laws follow
from proof-irrelevance.

@@

\subsection{Setoid Type Theory}

The setoid model presented in the previous section is peculiar in several
ways. One is that it can be presented in a very small type theoretic
metatheory. The other important aspect is that it is \emph{strict}, namely every
equation of the original syntax holds by definition in the semantics. One
advantage of strict models is that they can be turned into \emph{syntactic
  translations}, in which syntactic objects of the source theory are interpreted
as their counterparts in another \emph{target} theory. In the case of the setoid
model, this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on.

@@

\section{Universe of setoids}

@@

\subsection{Encoding of universes via IR and II}

@@

\subsection{A first definition of the setoid universe}

@@

\subsection{Inductive-inductive universe}

@@

\section{Indexed inductive universe (maybe?)}

@@

@@

\section{Further work}

>>>>

We begin by describing the metatheory that we will use throughout the paper, in
Section~\ref{sec:metatheory}. In Section~\ref{sec:setoid-model}, after briefly
recalling the abstract notion of models of Type Theory known as categories with
familes (CwF), we outline Altenkirch's setoid model of Type Theory in the
variant used in \cite{mpc19}. We then show how the model has been turned into
Setoid Type Theory in Section~\ref{sett}.

In Section~\ref{setoid-universe} we provide the actual construction of the
setoid universe in the setoid model. In Section~\ref{ir-universes} we first
recall the well known inductive-recursive definition of universes in Type
Theory, and the way these can be turned into a construction that does not
require induction-recursion.
%
We then provide, in Section~\ref{ir-setoid-universe}, a first complete
definition of the setoid universe using a quite complex form of
induction-recursion. This is followed by an alternative definition of the
universe in Section~\ref{ii-universe}, that does not rely on induction-recursion
but instead on infinitary induction-induction.
%
Further reductions of this inductive-inductive universe to even simpler forms,
in addition to other long term goals, are discussed in
Section~\ref{further-work}.

@@

This section describes $MLTT^{Prop}$, the metatheory in which the model
construction described in the following sections takes place. One aspect of the
metatheory that we pay attention to is it's minimality, as we want to keep it as
simple as possible. In particular, we want to avoid having to assume strong
definition schemas that go beyond inductive families.

@@

\section{Setoid model}\label{setoid-model}

@@

By \emph{setoid model} we intend a class of models of Type Theory where
contexts/closed types are interpreted as setoids, i.e. sets with an equivalence
relation, and dependent types are interpreted as dependent/indexed setoids.
%
A setoid model was first given for intensional Type Theory by M. Hofmann in
\cite{hofmann-ext-int}, in order to give a semantics to extensionality
principles such as function and propositional extensionality. Here we consider
a similar model construction due to Altenkirch \cite{setoid-model}.
%
The peculiarity of this model is that it is presented in a type theoretic and
intensional metatheory which includes a strict universe of propositions.
%
The use of strict propositions to define the setoid equivalence relations makes
this model \emph{strict}. A strict model is one where every equation holds
definitionally. The setoid model thus defined validates function extensionality,
a univalent universe of propositions, and quotient types. Therefore, it provides
a way to bootstrap and ``explain'' extensionality, since the model construction
effectively gives an implementation of various extensionality principles it in
terms of a small, entirely intensional type theory.

\subsection{Setoid model as a CwF}

As any model of Type Theory, the setoid model can be framed categorically as a
category with families (CwF) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:

@@

In our presentation of the setoid model, contexts are given by setoids, that is
types together with an equivalence relation.

\[
\infer{
  \begin{aligned}
    & | \GG | : \mType \\
    & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
    & \textsf{refl}\ \GG : (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
    & \textsf{sym} \ \GG : \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_0 \\
    & \textsf{trans} \ \GG : \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
    \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_2
    \to \GG^\sim\,\gamma_0\,\gamma_2
  \end{aligned}
}{
  \GG : \Con
}
\]

Types in a context $\GG$ are given by setoid morphisms from $\GG$ to the
(strict) category of setoids. The definition can be unfolded as follows.

\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & ... \\
    & \textsf{coe} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \textsf{coh} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \ (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

This definition of types in the setoid model is different from the one in [cite
  LICS 99 paper], but it is equivalent to it [cite]. The main difference here is
in the use of heterogeneous equality in the definition of types.

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.

@@

\subsection{Setoid Type Theory}\label{sett}

The setoid model presented in the previous section is peculiar in several
ways. One is that it can be presented in a very small type theoretic
metatheory. The other important aspect is that it is \emph{strict}, namely every
equation of the original syntax holds by definition in the semantics. One
advantage of strict models is that they can be turned into \emph{syntactic
  translations}, in which syntactic objects of the source theory are interpreted
as their counterparts in another \emph{target} theory. In the case of the setoid
model, this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on. \todo{maybe expand this?}

@@

\section{Universe of setoids}\label{setoid-universe}

@@

\subsection{Inductive-recursive universes in Type Theory}\label{ir-universes}

@@

\subsection{A first definition of the setoid universe}\label{ir-setoid-universe}

@@

\subsection{Inductive-inductive universe}\label{ii-universe}

@@

\section{Inductive universe (maybe?)}

@@

\section{Further work}\label{further-work}

@@
